Modal operators for a constructive account of well quasi-orders

### 2025-06-08 slides-types2025.lagda.tex
Final info slide
Some minor cleanups

  We then have:
  \begin{enumerate}
    \item $\mathsf{Well}\ind(X,{\le}) \Longleftrightarrow
    \necessary\mathsf{Well}\sinf(X,{\le})$. \formalized
    \item $(\possible\varphi) \Longrightarrow \varphi$,
    if $\varphi$ is a geometric implication (``$\forall\ldots\forall (\%
    \Rightarrow \%)$'', with no~$\forall$ nor~$\Rightarrow$
    in~$\%$).~\formalized
    \item $(\xpossible\varphi) \Longrightarrow \varphi$,
    if~$\varphi$ is a bounded first-order statement. \formalized
    \item For every inhabited type~$X$, $\xpossible\exists f : \NN
    \to X\_ \mathsf{Surjective}(f)$. \formalized
    \item $\possible \badbox{\textsc{lem}}$.
    (Baby Barr / Friedman's trick / nontrivial exit continuation)
  \end{enumerate}
\end{frame}}

\begin{frame}{The plain combinatorics of toposes}
  \begin{enumerate}\justifying
    \item Realize a generic gadget as some kind of limit of \hil{approximations}
    from the base universe. For the generic function $\NN \to X$:

@@

    \pause
    \item \hil{Reinterpret}, in a mechanical fashion, assertions about the generic
    gadget as assertions about its approximations~(\formalized). Be prepared to evolve
    approximations.

@@

    \begin{itemize}
    \item
    For a \hil{stage-dependent proposition}~$P : L \to \mathsf{Prop}$, $\nabla\,P\,\sigma$ expresses
    that no matter how~$\sigma$ will evolve over a time to a better
    approximation~$\tau$, eventually~$P\,\tau$ will hold.
  % later₂ : (a : X) → ((τ : X*) → a ∈ σ ⧺ τ → ∇ P (σ ⧺ τ)) → ∇ P σ
    \item We have the stage-dependent proposition ``$f_0\,n = x$'',
    where~$f_0$ is the generic function.
    \item That~$f_0$ is defined on input~$n$ can be expressed
    as~$\nabla\,P\,[\,]$ where~$P\,\sigma \defeqv (\mathsf{length}\,\sigma > n)$.\end{itemize}}
    \medskip
    \pause

@@

\begin{frame}{Increasing subsequences as convenient fictions}
  \justifying
  Let~$X$ be a quasi-order.
  Let~$B : \mathsf{List}\,X \to \mathsf{Prop}$ be a monotone predicate.

@@

  \begin{block}{Constructive reimagination}
    \textbf{Thm.} If~$X$ is well\ind and if~$\nabla^\nnearrow\,B\,[\,]$,
    then~$\nabla\,B\,[\,]$. \formalized
  \end{block}
  \vspace*{-0.4em}
  {\small\emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence (in an arbitrary topos).
  \emph{Somewhere,} \textsc{lem} holds. \emph{There}~$X$
  is still well\sinf, so that we have an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
  By assumption, this subsequence has a finite prefix validating~$B$. This
  prefix is part of a prefix of the original sequence~$\alpha$. Hence~$\alpha$
  has a prefix validating~$B$ by monotonicity \emph{there}. So
  \emph{somewhere} there is a finite prefix validating~$B$. Thus there
  actually is a finite prefix validating~$B$. \qed\par}
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Maximal ideals as convenient fictions}
  \justifying
  Let~$A$ be a commutative ring with unit.

@@

  Note: We can use~$\nabla$ even without a proper metatheoretic backing.
\end{frame}
\backupstart

>>>>

  We then have:
  \begin{enumerate}
    \item $\mathsf{Well}\ind(X,{\le}) \Longleftrightarrow
    \necessary\mathsf{Well}\sinf(X,{\le})$. \formalized
    \item $(\possible\varphi) \Longleftrightarrow \varphi$,
    if $\varphi$ is a geometric implication (``$\forall\ldots\forall (\%
    \Rightarrow \%)$'', with no~$\forall$ nor~$\Rightarrow$
    in~$\%$).~\formalized
    \item $(\xpossible\varphi) \Longleftrightarrow \varphi$,
    if~$\varphi$ is a bounded first-order statement. \formalized
    \item For every inhabited type~$X$, $\xpossible\necessary
    \mathsf{Countable}(X)$, where~$\mathsf{Countable}(X) \defeqv (\exists f : \NN
    \to X\_ \mathsf{Surjective}(f))$. \formalized

    So being countable is a \hil{button}.
    \item $\possible \badbox{\textsc{lem}}$.
    (Baby Barr / Friedman's trick / nontrivial exit continuation)

    In fact, \badbox{\textsc{lem}} is a \hil{switch}:
    $\necessary\bigl((\possible\badbox{\textsc{lem}}) \wedge
    (\possible\neg\badbox{\textsc{lem}})\bigr)$

    \item $\badbox{\textsc{zorn}} \Rightarrow
    \necessary\possible\badbox{\textsc{ac}}$. (Great Barr)
  \end{enumerate}
\end{frame}}

\begin{frame}{The plain combinatorics of toposes}
  \begin{enumerate}\justifying
    \item Realize a generic gadget as some kind of limit of \hil{approximations}
    from the base universe. For the generic function $f_0 : \NN \to X$:

@@

    \pause
    \item \hil{Reinterpret}, in a mechanical fashion, assertions about the generic
    gadget as assertions about its approximations~(\formalized).

    {\small\begin{itemize}
    \item We have the \hil{stage-dependent proposition} ``$f_0\,n = x$'',
    a certain function $L \to \mathsf{Prop}$: $\lambda \sigma.\
    (\mathsf{lookupMaybe}\,\sigma\,n = \mathsf{just}\,x)$\end{itemize}}
    \medskip

    \pause
    \item Be prepared to evolve approximations (\formalized).

@@

    \begin{itemize}
    \item
    For a stage-dependent proposition~$P : L \to \mathsf{Prop}$, $\nabla\,P\,\sigma$ expresses
    that no matter how~$\sigma$ will evolve over a time to a better
    approximation~$\tau$, eventually~$P\,\tau$ will hold.
  % later₂ : (a : X) → ((τ : X*) → a ∈ σ ⧺ τ → ∇ P (σ ⧺ τ)) → ∇ P σ
    \item That~$f_0$ is defined on input~$n$ can be expressed
    as~$\nabla\,P\,[\,]$ where~$P\,\sigma \defeqv (\mathsf{length}\,\sigma > n)$.\end{itemize}}
    \medskip
    \pause

@@

  Note: We can use~$\nabla$ even without a proper metatheoretic backing.
\end{frame}

{\renewcommand{\insertframeextra}{a}
\begin{frame}{Increasing subsequences as convenient fictions}
  \justifying
  Let~$X$ be a quasi-order.
  Let~$B : \mathsf{List}\,X \to \mathsf{Prop}$ be a monotone predicate.

@@

  \begin{block}{Constructive reimagination}
    \textbf{Thm.} If~$X$ is well\ind and if~$\nabla^\nnearrow\,B\,[\,]$,
    then~$\nabla\,B\,[\,]$. \formalized
  \end{block}
  \vspace*{-0.4em}
  {\small\emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence (in an arbitrary topos).
  \emph{Somewhere,} \textsc{lem} holds. \emph{There}~$X$
  is still well\sinf, so that we have an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
  By assumption, this subsequence has a finite prefix validating~$B$. This
  prefix is part of a prefix of the original sequence~$\alpha$. Hence~$\alpha$
  has a prefix validating~$B$ by monotonicity \emph{there}. So
  \emph{somewhere} there is a finite prefix validating~$B$. Thus there
  actually is a finite prefix validating~$B$. \qed\par}
\end{frame}}

\addtocounter{framenumber}{-1}
{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\renewcommand{\insertframeextra}{b}
\begin{frame}{Maximal ideals as convenient fictions}
  \justifying
  Let~$A$ be a commutative ring with unit.

@@

\backupstart

\begin{frame}[plain]
  \begin{columns}[c]
    \begin{column}{0.45\textwidth}
      \includegraphics[width=0.95\textwidth]{lets-play-agda}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=0.8\textwidth]{ai-2025}

      \vspace*{-1.5cm}\hspace*{2cm}
      \includegraphics[width=0.8\textwidth]{cirm-2025}

      \vspace*{1cm}
    \end{column}
  \end{columns}
\end{frame}
